Replicating Blair and Sambanis: Forecasting Civil Wars - JCR 2020

### 2020-04-29 issues.Rmd
update issues doc

- Improperly tuning of models
  - RM: In a twitter DM, I asked Blair to clarify -- "[I] was wondering how you all came to your tuning procedure for the RF model hyper-parameters?" He responded "mostly trial and error, honestly. ... Trial and error was on early data. Forecasts were for much later data"
- Using a regression framework in a classification problem
- Their train/test approach 
  - 5-year forecasts for sep plots and AUC scores)
- Rounding of AUC/ROC scores
- Lack of yearly test forecasts in favor of a single 5-year test forecast
- Table 4! 
  - There are only 20 onsets from 2001 to 2015 But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)
- Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework
- The lack of procedures to account for rare events in an RF model

@@

1. Full replication
    - Can we get their results
    - RM: maybe I can go through their code and bring it "up-to-date" i.e. make it tidy... 
    - Figure out how they built table 4 (It looks like this is done in STATA)
    - Check to see if it matter that they are using regression for a classification problem
2. Address tuning 
    - Does properly tuning the CAMEO model improve its performance?
3. Address AUC/ROC smoothing... 
    - How drastic is the change; does it change their conclusions? 
4. Run a yearly test forecast -- as is the current standard in TSCS forecasts (Is it the current standard?) 
5.

>>>>

1. Using AUC for smoothed ROC curves. Does the interpretation of Table 1, and superiority of the escalation model, depend on the choice to use smoothed ROC curves?

2. In Table 4, are the forecasts evaluated against incidence or onset? The data from 2001 to 2015 have 20 civil war onsets in total, yet Table 4 appears to show 15 and 16 positive events for 2016 only. If this is indeed incidence, but the original models and forecasts were for onset, then this is a incorrect assessment of the forecasts' accuracy.

- Table 4! 
  - There are only 20 onsets from 2001 to 2015 But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)

3. Unorthodox RF hyperparameter choices that may be driving the OOS test prediction results. There are two issues here. One is that the RF models are so unorthodox. I'm not sure that this in itself is a problem. But the second potential problem is that the hyperparameter values may not have been tuned in a way that works the best for the non-escalation model, and especially the CAMEO model with 1,100+ features.

- Improperly tuning of models
  - RM: In a twitter DM, I asked Blair to clarify -- "[I] was wondering how you all came to your tuning procedure for the RF model hyper-parameters?" He responded "mostly trial and error, honestly. ... Trial and error was on early data. Forecasts were for much later data"
- Using a regression framework in a classification problem
- Rounding of AUC/ROC scores
    + AB: smoothing or rounding? I think rounding is ok. 
- Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework
    + It's hacky but it works. The resulting forecasts are within the 0-1 interval.

The table below shows the randomForest package default hyperparameter values for a binary classification problem like the one at hand and compares them to the B&S base settings. Some hyperparameters have heuristics that determine the default value based on characteristics of the input data; we note these in the second column and what the realized default setting would be for the basic Escalation model (first row in Table 1 in the paper) in the 3rd column. The last column has the settings B&S use.

|Hyperparameter|Default heuristic|Default values (Escalation)|B&S value|
|--------------|-----------------|--------------------------|---------|
|type          |                 | classification           | regression|
|ntree         |                 | 500                      | 100,000 |
|mtry          |`floor(sqrt(ncol(x)))`|3|3|
|replace       |                 | true | false |
|sampsize | `nrow(x)` if replace, else `ceiling(.632*nrow(x))` | 11,869 | 100 |
|nodesize | 1 for classification | 1 | 1 |
|maxnodes | | null | 5 |

It is worth noting that commonly the ntree, mtry, and nodesize parameters are the main parameters tuned; of these three in the B&S specification only 1, the ntree parameter, deviates from the default settings.

In any case, there is a stark contrast in the default RF settings and the way B&S use the RF models. The default approach is to train a relatively small number (ntree; 500) of classification trees, but where each tree is fairly big in that it is trained on data that has the same number of rows as the training data, albeit sampled with replacement (replace is true; sampsize is 11,869), and allowed to grow fairly deep (this is governed by nodesize (1), which is the minimum size a terminal node must have). In contrast, B&S grow very extensive forests with a large number of trees (100,000 compared to 500), but each tree is very small and shallow; only a 100 rows are sampled from the training data for each tree, and the trees are constrained to at most 5 terminal nodes (maxnodes).

This approach only works due to the other unorthodox choice, which is to use regression, not classificaton, trees. Trying to use classification trees with the other parameter settings in fact does not work at all because it is almost guaranteed that a sample of 100 from the 11,869 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample. As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:

```
Warning message:
In randomForest.default(y = as.integer(train_df$incidence_civil_ns_plus1 ==  :
  The response has five or fewer unique values.  Are you sure you want to do regression?
```

While unorthodox, the approach does work. It produces predictions that are within the 0 to 1 interval. The concern is that the settings work well only for the escalation model, and only for the particular test set at hand. Specifically we wonder if the CAMEO model with 1,100+ features, compared to 10 or less for the other specifications, would perform better with a more explicit tuning procedure.

The table below shows cross-validated training and OOS test AUC-ROC for the original 1 month escalation model and a modified version which uses more conventional random forest parameter settings (classification, 5000 trees, and default settings otherwise). The first and second columns are average and SD AUC-ROC obtained through repeated cross-validation on the training data set. Because there are only 9 positive cases in the whole training data, and thus to ensure that any given split will include at least 1 positive case in each data partition, we use 2-fold CV, i.e. splitting the original training data into equally-sized new training and validation sets. This is repeated 21 times for a total of 42 OOS performance samples for each model. The last column shows the OOS performance on the original test set. The first value in this column corresponds to the base escalation model results reported Table 1 in the B&S paper when *not* using smoothed ROC curves.

```{r}
tbl <- structure(list(Model = c("Escalation, 1mo", "Modified Escalation, 1mo"
), Avg_CV_ROC_AUC = c(0.698772142673124, 0.643982732304452), 
    SD_CV_ROC_AUC = c(0.136834250535024, 0.104102374345325), 
    Test_ROC_AUC = c(0.786872878159185, 0.583391239320482)), row.names = c(NA, 
-2L), class = c("tbl_df", "tbl", "data.frame"))

knitr::kable(tbl, digits = 2)
```

Two thigs stand out:

- The escalation model has better performance in the test data than in the training CV samples. This is consistent with hyperparameter values that were (over-)fit to the test data. 
- Although the original escalation model outperforms the modified model in the training CV resamples (p < 0.05), the difference in means is much less pronounced than in the test data.

4. Design choices

Rick, I think the points below are more subjective choices. The first two points above are IMO potentially objectively incorrect technical errors that could undermine the original B&S findings. The 3rd point about the RF models is somewhere between objective error and subjective choice. I'm not sure. But the points below I would say are subjective choices that could alter their findings and that I think are justifiable criticisms, but I think it'll be easier for someone to push back on this if they wanted to.

- Their train/test approach 
  - 5-year forecasts for sep plots and AUC scores)
- Lack of yearly test forecasts in favor of a single 5-year test forecast
- The lack of procedures to account for rare events in an RF model

@@

1. Full replication
    - Can we get their results
    - RM: maybe I can go through their code and bring it "up-to-date" i.e. make it tidy... 
        + AB: that sounds like a lot of work
    - Figure out how they built table 4 (It looks like this is done in STATA)
    - Check to see if it matter that they are using regression for a classification problem
        + AB: i tried switching the models as they have them just from regression to classification and it breaks because the data samples end up being all 0's. 
2. Address tuning 
    - Does properly tuning the CAMEO model improve its performance?
3. Address AUC/ROC smoothing... 
    - How drastic is the change; does it change their conclusions? 
4. Run a yearly test forecast -- as is the current standard in TSCS forecasts (Is it the current standard?) 
5.
